Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-24 ind_files/06_nontechnical_discussion.tex
Make changes as suggested by reviewers

Stochastic models in biology often exhibit behaviors that are markedly different from their deterministic limits~\citep{jafarpour_noise-induced_2017,boettiger_noise_2018,jhawar_noise-induced_2020,coomer_noise_2022}. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems~\citep{black_stochastic_2012,hastings_transients_2004,shoemaker_integrating_2020,schreiber_does_2022}. In the context of population biology, the finiteness of populations is an important source of stochasticity that can lead to behavior not captured in corresponding infinite population limits~\citep{black_stochastic_2012,rogers_demographic_2012,debarre_evolutionary_2016,delong_stochasticity_2023}. When actually incorporating this stochasticity into eco-evolutionary theory, it is important to build up the stochastic theory from first principles instead of adding noise to existing deterministic models in an ad-hoc fashion, since the latter procedure can very easily lead to \textcolor{red}{\st{inconsistent models}} \textcolor{blue}{ill-behaved models with un-biological properties}~\citep{strang_how_2019}. Indeed, several theorists have called for a `bottom-up' reformulation of eco-evolutionary dynamics from the first principles of stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic~\citep{metcalf_why_2007,lambert_population_2010, geritz_mathematical_2012,doebeli_towards_2017}.

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. \textcolor{red}{\st{In particular, my equations generalize the unifying formalism described in Lion, 2018 to finite, fluctuating populations - taking $K \to \infty$ in equations (5.1), (5.2), and (5.9) recover equations (6), (11), and (14) in Lion, 2018 respectively as their infinite population limits}} \textcolor{blue}{The equations I derive deal with biologically important quantities in a `coarse-grained' manner, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of~\cite{queller_fundamental_2017}) and present a `unifying perspective' (in the sense of~\cite{lion_theoretical_2018}) for the dynamics of finite populations.} While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

\textcolor{blue}{Populations consisting mostly of types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will experience a reduced magnitude of noise-induced selection acting in the system compared to Populations consisting mostly of types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by evolution through classical selection and noise-induced selection respectively (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual). This is why the direction of deterministic selection is specifically reversed to favor mutualists in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments~\citep{sanchez_feedback_2013} and finite population IbMs~\citep{houchmandzadeh_selection_2012,,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019} of social evolution despite infinite population models predicting their extinction}.

\textcolor{blue}{The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~\citep{mcleod_social_2019}. In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of }\eqref{nD_stochastic_Price} \textcolor{blue}{becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of noise-induced selection.}

@@

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. Since noise-induced selection is directional, the seemingly small systematic deviations from neutrality exhibited by these models can have significant consequences over long evolutionary times (For example, see~\cite{veller_drift-induced_2017}). My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print~\citep{kuosmanen_turnover_2022} \textcolor{red}{\st{that independently arrives at my equations for type frequencies (equation (5.1)) and the change of mean fitness and turnover in the population (equations (5.7) and (5.8) respectively) using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration (Kuosmanen et al., 2022).}}.

On the practical side, the existence of noise-induced selection implies that simulation studies working with evolutionary individual-based or agent-based models should be careful about whether interaction effects are incorporated into birth rates or death rates since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection, thus potentially biasing results~\citep{mcleod_social_2019,kuosmanen_turnover_2022}. My results also indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction/inference of future trajectories of the relative abundance of a species (or phenotype, allele, etc.) from empirical data even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i} - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates. In contrast, the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i} + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates). This discrepancy can lead to systematic deviations from the trajectories predicted using the growth rate alone, especially in smaller populations. \textcolor{blue}{This deviation should be empirically measurable. Recent improvements in statistical inference methods for birth and death rates suggest that we should also be able to estimate the birth and death rates of real populations~\citep{huynh_inferring_2023} and thus the quantitative predictions of our SDEs, providing a direct empirical test of theoretical predictions. However, since the strength of noise-induced selection scales inversely with the total population size, we can expect such deviations to quickly become too small to detect in medium to large populations, though this is purely due to limits on the precision of experimental measurements rather than due to any theoretical limitations.}

Being mindful of noise-induced selection is also important for applied fields that regularly deal with manipulating small populations, such as conservation and population management. For example, when trying to increase the population of a hypothetical desired species in a multi-species community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which is favored by noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community (See Appendix \ref{App_deviation_from_neutrality} for a simple example that illustrates this). If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: \textcolor{blue}{This study shows that in their model,} due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

@@

\textcolor{blue}{~\cite{lion_theoretical_2018} has recently proposed a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework using `feedback' as a `unifying perspective'. Our equations generalize the unifying framework described in~\cite{lion_theoretical_2018} to finite, fluctuating populations --- taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - equation \eqref{nD_stochastic_RM} is the fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). If we assume that the quantity $f$ follows a Gaussian distribution, then the mean and variance completely characterize the distribution,  and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system.}

\textcolor{blue}{The equivalent of our stochastic equations has recently been derived for quantitative traits from a very different starting point using the theory of measure-valued branching processes~\citep{week_white_2021} --- Equations (21b) and (21c) in \cite{week_white_2021} are exactly the $m \to \infty$ version of our equations for changes in the mean value of a type-level quantity and changes in the variance of a type-level quantity respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. The methods used by these authors is exactly complementary to the field equations we formulate in Chapter~\ref{chap_infD_processes} (See section~\ref{sec_disc_field_eqns} for a more detailed technical discussion). A recent preprint has also independently arrived at our equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover}) in the context of life-history evolution using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.}

\textcolor{blue}{A recent study showed that in a very broad class of density-dependent competition models, model-specific details can be ignored in favor of a universal `coarse-grained' description that only includes a small number of very generally defined quantities~\citep{mazzolini_universality_2022}. Among the generic predictions of this general description was the presence of noise-induced selection in general competition models. Our work generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. Our results also show that equations like the replicator-mutator equation and Price equation are asymptotically `universal', in the sense that an arbitrary density-dependent birth-death process with functional forms given by \eqref{nD_functional_forms_for_replicator} will satisfy the replicator-mutator or Price equation as we move to infinitely large populations ($K \to \infty$). Thus, these equations provide coarse-grained `fundamental theorems' that are always satisfied for biological populations that are sufficiently large.}

\textcolor{blue}{Rice has proposed a stochastic version of the Price equation in a series of papers~\citep{rice_stochastic_2008,rice_evolution_2009,rice_universal_2020} that follow the classic derivation by Price himself~\citep{rice_evolutionary_2004}. Like the original Price equation, these equations are formulated in a very general manner that relates the phenotypic change between two given populations. As such, they are more general than my equations, but, just like the original Price equation, are dynamically insufficient and cannot be phrased in the language of dynamical systems (to the best of my understanding). They are thus the true stochastic analog of the original Price equation, whereas my version (equation \eqref{nD_stochastic_Price}) is the analog of~\posscite{lion_theoretical_2018} version of the Price equation in a continuous time, dynamically sufficient setting. The two formulations are complementary --- Rice's equations (and the original Price equation) provide a very general description for partitioning phenotypic change between two time points, whereas my approach (and ~\posscite{lion_theoretical_2018} version of the Price equation) provides equations that are slightly less general, but are formulated in a continuous time, dynamically sufficient, predictive manner. Rice's derivations also treat fitness as fundamental, whereas I derive suitable notions of fitness and turnover from demographic first principles. As a consequence, the `extra' stochastic term corresponding to noise-induced selection that appears in my equations fundamentally emerges from the stochasticity of the underlying births and deaths of organisms and is thus of ecological/demographic origin, whereas the `extra' stochastic term in Rice's equations emerges from the stochasticity of fitness\footnote{defined here as the number of offspring produced --- no deaths involved~\citep{rice_stochastic_2008,rice_evolution_2009,rice_universal_2020}}alone and, to the best of my knowledge, does \emph{not} correspond to the same effect I identify in this thesis. Lastly, whereas Rice's equations are exact, my equations are derived using a system-size expansion (see part \ref{part_theory}) and thus are, strictly, speaking only an approximate statistical description of the dynamics (though these approximations are usually very good for even moderate values of $K$~\citep{black_stochastic_2012,cianci_wkb_2015})}

>>>>

Stochastic models in biology often exhibit behaviors that are markedly different from their deterministic limits~\citep{jafarpour_noise-induced_2017,boettiger_noise_2018,jhawar_noise-induced_2020,coomer_noise_2022}. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems~\citep{black_stochastic_2012,hastings_transients_2004,shoemaker_integrating_2020,schreiber_does_2022}. In the context of population biology, the finiteness of populations is an important source of stochasticity that can lead to behavior not captured in corresponding infinite population limits~\citep{black_stochastic_2012,rogers_demographic_2012,debarre_evolutionary_2016,delong_stochasticity_2023}. When actually incorporating this stochasticity into eco-evolutionary theory, it is important to build up the stochastic theory from first principles instead of adding noise to existing deterministic models in an ad-hoc fashion, since the latter procedure can very easily lead to ill-behaved models with un-biological properties~\citep{strang_how_2019}. Indeed, several theorists have called for a `bottom-up' reformulation of eco-evolutionary dynamics from the first principles of stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic~\citep{metcalf_why_2007,lambert_population_2010, geritz_mathematical_2012,doebeli_towards_2017}.

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. The equations I derive deal with biologically important quantities in a `coarse-grained' manner, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of~\cite{queller_fundamental_2017}) and present a `unifying perspective' (in the sense of~\cite{lion_theoretical_2018}) for the dynamics of finite populations. While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

Populations consisting mostly of types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will experience a reduced magnitude of noise-induced selection acting in the system compared to Populations consisting mostly of types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by evolution through classical selection and noise-induced selection respectively (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual). This is why the direction of deterministic selection is specifically reversed to favor mutualists in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments~\citep{sanchez_feedback_2013} and finite population IbMs~\citep{houchmandzadeh_selection_2012,,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019} of social evolution despite infinite population models predicting their extinction.

The fact that total population size controls the strength of noise-induced selection also explains why cooperation is favored in the early transient period of population growth~\citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size --- In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favoring cooperation. The fact that the entire term scales inversely with the total population size $KN_K(t)$ suggests that the effect of this force is weak for large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, timescale separation, or a weak selection + weak mutation limit~\citep{mcleod_social_2019}. In all three of these cases, the effects of classical selection vanish, and thus the first term on the RHS of \eqref{nD_stochastic_Price} becomes identically 0. It therefore no longer contributes to the trait frequency dynamics, allowing us to see the (otherwise weak) contributions of noise-induced selection.

@@

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. Since noise-induced selection is directional, the seemingly small systematic deviations from neutrality exhibited by these models can have significant consequences over long evolutionary times (For example, see~\cite{veller_drift-induced_2017}). My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print~\citep{kuosmanen_turnover_2022}.

On the practical side, the existence of noise-induced selection implies that simulation studies working with evolutionary individual-based or agent-based models should be careful about whether interaction effects are incorporated into birth rates or death rates since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection, thus potentially biasing results~\citep{mcleod_social_2019,kuosmanen_turnover_2022}. My results also indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction/inference of future trajectories of the relative abundance of a species (or phenotype, allele, etc.) from empirical data even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i} - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates. In contrast, the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i} + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates). This discrepancy can lead to systematic deviations from the trajectories predicted using the growth rate alone, especially in smaller populations. This deviation should be empirically measurable. Recent improvements in statistical inference methods for birth and death rates suggest that we should also be able to estimate the birth and death rates of real populations~\citep{huynh_inferring_2023} and thus the quantitative predictions of our SDEs, providing a direct empirical test of theoretical predictions. However, since the strength of noise-induced selection scales inversely with the total population size, we can expect such deviations to quickly become too small to detect in medium to large populations, though this is purely due to limits on the precision of experimental measurements rather than due to any theoretical limitations.

Being mindful of noise-induced selection is also important for applied fields that regularly deal with manipulating small populations, such as conservation and population management. For example, when trying to increase the population of a hypothetical desired species in a multi-species community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which is favored by noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community (See Appendix \ref{App_deviation_from_neutrality} for a simple example that illustrates this). If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species. Indeed, numerical investigations of `burst-death' dynamics in fluctuating viral populations show that increasing Malthusian fitness by boosting the survival rate (\emph{i.e.} reducing the death rate) leads to a greater increase in fixation probability than if an exactly equivalent increase in Malthusian fitness is achieved via increasing the burst rate (\emph{i.e.} increasing the birth rate) of viral particles~\citep{alexander_fixation_2008}.  A recent finite population birth-death model for cancer treatment provides another concrete example of the consequences of the asymmetry between changing birth rates and death rates: This study shows that in their model, due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells, with obvious implications for optimal treatment strategies~\citep{raatz_promoting_2023}.

@@

~\cite{lion_theoretical_2018} has recently proposed a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework using `feedback' as a `unifying perspective'. Our equations generalize the unifying framework described in~\cite{lion_theoretical_2018} to finite, fluctuating populations --- taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - equation \eqref{nD_stochastic_RM} is the fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). If we assume that the quantity $f$ follows a Gaussian distribution, then the mean and variance completely characterize the distribution,  and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system.

The equivalent of our stochastic equations has recently been derived for quantitative traits from a very different starting point using the theory of measure-valued branching processes~\citep{week_white_2021} --- Equations (21b) and (21c) in \cite{week_white_2021} are exactly the $m \to \infty$ version of our equations for changes in the mean value of a type-level quantity and changes in the variance of a type-level quantity respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. The methods used by these authors is exactly complementary to the field equations we formulate in Chapter~\ref{chap_infD_processes} (See section~\ref{sec_disc_field_eqns} for a more detailed technical discussion). A recent preprint has also independently arrived at our equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover}) in the context of life-history evolution using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.

A recent study showed that in a very broad class of density-dependent competition models, model-specific details can be ignored in favor of a universal `coarse-grained' description that only includes a small number of very generally defined quantities~\citep{mazzolini_universality_2022}. Among the generic predictions of this general description was the presence of noise-induced selection in general competition models. Our work generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. Our results also show that equations like the replicator-mutator equation and Price equation are asymptotically `universal', in the sense that an arbitrary density-dependent birth-death process with functional forms given by \eqref{nD_functional_forms_for_replicator} will satisfy the replicator-mutator or Price equation as we move to infinitely large populations ($K \to \infty$). Thus, these equations provide coarse-grained `fundamental theorems' that are always satisfied for biological populations that are sufficiently large.

Rice has proposed a stochastic version of the Price equation in a series of papers~\citep{rice_stochastic_2008,rice_evolution_2009,rice_universal_2020} that follow the classic derivation by Price himself~\citep{rice_evolutionary_2004}. Like the original Price equation, these equations are formulated in a very general manner that relates the phenotypic change between two given populations. As such, they are more general than my equations, but, just like the original Price equation, are dynamically insufficient and cannot be phrased in the language of dynamical systems (to the best of my understanding). They are thus the true stochastic analog of the original Price equation, whereas my version (equation \eqref{nD_stochastic_Price}) is the analog of~\posscite{lion_theoretical_2018} version of the Price equation in a continuous time, dynamically sufficient setting. The two formulations are complementary --- Rice's equations (and the original Price equation) provide a very general description for partitioning phenotypic change between two time points, whereas my approach (and ~\posscite{lion_theoretical_2018} version of the Price equation) provides equations that are slightly less general, but are formulated in a continuous time, dynamically sufficient, predictive manner. Rice's derivations also treat fitness as fundamental, whereas I derive suitable notions of fitness and turnover from demographic first principles. As a consequence, the `extra' stochastic term corresponding to noise-induced selection that appears in my equations fundamentally emerges from the stochasticity of the underlying births and deaths of organisms and is thus of ecological/demographic origin, whereas the `extra' stochastic term in Rice's equations emerges from the stochasticity of fitness\footnote{defined here as the number of offspring produced --- no deaths involved~\citep{rice_stochastic_2008,rice_evolution_2009,rice_universal_2020}}alone and, to the best of my knowledge, does \emph{not} correspond to the same effect I identify in this thesis. Lastly, whereas Rice's equations are exact, my equations are derived using a system-size expansion (see part \ref{part_theory}) and thus are, strictly, speaking only an approximate statistical description of the dynamics (though these approximations are usually very good for even moderate values of $K$~\citep{black_stochastic_2012,cianci_wkb_2015}).
